Causal Analysis Paper

### 2022-06-08 PaperAnalysis.Rmd
First edits to the final version

Abortion law and the negative consequences of the lack of safe abortions are highly contentious policy issues that have been making headlines this year. Further, the Sustainable Development Goal indicator 3.1 of the United Nations to reduce Maternal Death to less than 70 per 100000 life birth by 2030 (https://www.who.int/data/gho/indicator-metadata-registry/imr-details/26) is relying on access to safe and legal abortions. In the paper we are analysing the authors are testing whether and if so to what extent abortion law as a proxy for access to safe abortions causally effects maternal mortality rates in over 160 countries around the globe. Establishing a causal relation could strengthen the implementation of abortion liberalisation policy. This is especially needed in a time where several states already withdrew or are withdrawing access to safe abortions on many grounds such as Poland (https://www.amnesty.org/en/latest/news/2021/10/poland-a-year-on-abortion-ruling-harms-women/) and the USA (https://www.euronews.com/2022/05/03/us-supreme-court-moving-to-overturn-key-abortion-rights-law-says-leaked-report). While on the other hand there are developments like for instance in France to expand abortion rights (https://www.hrw.org/news/2022/03/01/france-expands-abortion-access-two-key-moves) in 2022. So does abortion law flexibility cause maternal mortality? We will be taking a close critical look at a paper that tries to establish exactly that and also run our own data analysis to address this question.

@@

Sensitivity Analysis

@@

### Missing data

Hernan et al. are outlining in their paper why data is missing, to restrict the analysis on countries where data are complete would bias it. Countries with missing data on female literacy are still included {@10.2307/20485961, p. 5}. Female primary completion is only available for 50 percent of countries and female literacy only for 10 percent. This missingness of data introduces selection bias to this specific analysis the auhtors did in the original paper. As the likely selection bias in this study cannot be avoided due to the limitations on data availability it needs to be corrected if possible, in the analysis {@10.2307/20485961, p. 7} this could be achieved through generalization of inverse probability weighting. For this however we would need to categorize developing countries in one group where some have recorded female literacy which then gets multiplied by the number of countries not considered due to missing data. Inverse probability weighting would work as the score for female literacy is not zero for the countries where it is not recorded.  However, a lot of assumptions would need to be made to apply this method. The assumption that the filler-country would be close enough to the unobserved countries is by nature untestable and therefor it is a strong assumption.

@@

[move Angelas equation here?]

@@

G-separation calculation has two steps: 
1. Fitting a Q-model, that is, a traditional regression of the outcome on the treatment and  relevant covariates.  In our case, we used the traditional regression fitted in the original paper as Q model:

$$
MMR=\beta_0 + \beta_1 L+\beta_2 GDP +\beta_3 P+ \beta_4 C
$$
In the formula, P and C respectively account for 3 year periods and the country specific effects. 
2. The Q-model is used to predict counterfactual outcomes for all the observations under both treatment settings (L=1 for flexible abortion laws and L=0 if not). More specifically, we plugged both treatment regimes in the the regression fit and obtained two predicted outcomes: $Y_{L=0}=Y_0$ for the unexposed and $Y_{L=1}=Y_1$ for the exposed. Both outcomes are then used to calculate the average $Y_1-Y_0$, which corresponds to the ATE.

@@

Furthermore, G-separation allows for the calculation of both ATT and ATU by restricting the ATE estimation to the treated and the untreated respectively[@Wang].

@@

* The updated dataset
Although the datasets used in the original paper were hosted in public websites, the currently available version is not exactly the same. Since both the abortion policies and the maternal mortality sets are now hosted in interactive webpages, it was hard to access the aggregated and historical data from both abortion policies and maternal mortality rates between 1985-2013. Despite these restrictions, we collected data on all the relevant variables from 2000 to 2017. The maternal mortality data was accessed in the World Bank Data repository[@MMR], and the global abortion policies were found in the UN Department for Economic and Social Affairs archive [@WorldAbortionPolicies].

@@

* Regression ATE
In the first part of our exploration, we followed the methodology of the original paper and used fixed effects linear regression models to estimate the relationship between abortion law flexibility and MMR. We controlled for GDP per capita, time in three-year intervals, and country-specific effects through a factor country-name variable. According to our results, flexible abortion laws reduce maternal deaths by 57 per 100.000 live births [ADD TABLE]. This result is close enough to the original result, which found that maternal deaths reduce by 45 per 100,000 live births (95% CI: -64, -26) with flexible abortion laws. 
We also conducted added "female primary education completion" as a control variable, and found very similar result: the effect of abortion law flexibility on maternal mortality appears to be a reduction in the mortality rate of 53 deaths per 100.000 live births.

* Heterogeinity of effects

@@

* G-separation ATE, ATU, ATT

* Unmeasured confounding

@@

# github link to our analysis project: https://github.com/aduarte56/AbortionLaws

@@

@@

Helena Bakic : Empirical estimation,Identification, Heterogeinity of effects, Analysis
Anna-Lisa Wirth (213286@mds.hertie-schoo.org) : Background, Unmeasured Confounding, Missing Data, Sensitivity Analysis, Analysis and Conclusion 
Ángela Duarte Pardo : Data Cleaning, G-separation, Methodology, Modelling and Analysis

@@

## Datasets
### Origital paper datasets
-   UN (2013). Population division. World Abortion policies 2013.  [Accessible here](https://www.un.org/en/development/desa/population/publications/policy/world-abortion-policies-2013.asp)

### Updated datasets
- WHO. (2022). Global Abortion Policies Worldwide. [Accessible here](https://abortion-policies.srhr.org/)

- WHO. (2022). Maternal Mortality Database.[Accessible here:](https://platform.who.int/mortality/themes/theme-details/topics/topic-details/MDB/maternal-conditions)

### Other Data sets

-	OSF. (2020). Global abortion Incidence (https://osf.io/6t4eh/)

- - WHO. (2022). Global Abortion Laws. [Accessible here](https://abortion-policies.srhr.org/)

-	Health system. (https://apps.who.int/nha/database), hospital beds (https://data.worldbank.org/indicator/SH.MED.BEDS.ZS)
-	Economic: GPD, human capital (education, as in Jensen paper), poverty (https://data.worldbank.org/indicator/SI.POV.NAHC), GINI (https://data.worldbank.org/indicator/SI.POV.GINI)

- life expectancy at birth (https://data.worldbank.org/indicator/SP.DYN.LE00.IN)

###Text

>>>>

Abortion law and the negative consequences of the lack of safe abortions are highly contentious policy issues that have been making headlines this year. Further, the [Sustainable Development Goal indicator 3.1](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/26) of the United Nations to reduce Maternal Death to less than 70 per 100000 life birth by 2030  is relying on access to safe and legal abortions. In the paper we are analysing the authors are testing whether and if so to what extent abortion law as a proxy for access to safe abortions causally effects maternal mortality rates in over 160 countries around the globe. Establishing a causal relation could strengthen the implementation of abortion liberalisation policy. This is especially needed in a time where several states already withdrew or are withdrawing access to safe abortions on many grounds such as [Poland](https://www.amnesty.org/en/latest/news/2021/10/poland-a-year-on-abortion-ruling-harms-women/) and the [USA](https://www.euronews.com/2022/05/03/us-supreme-court-moving-to-overturn-key-abortion-rights-law-says-leaked-report). While on the other hand there are developments like for instance in [France](https://www.hrw.org/news/2022/03/01/france-expands-abortion-access-two-key-moves) to expand abortion rights in 2022. So does abortion law flexibility cause maternal mortality? We will be taking a close critical look at a paper that tries to establish exactly that and also run our own data analysis to address this question.

@@

*Sensitivity Analysis*

@@

*Missing data*
Hernan et al. are outlining in their paper why data is missing, to restrict the analysis on countries where data are complete would bias it. Countries with missing data on female literacy are still included {@10.2307/20485961, p. 5}. Female primary completion is only available for 50 percent of countries and female literacy only for 10 percent. This missingness of data introduces selection bias to this specific analysis the auhtors did in the original paper. As the likely selection bias in this study cannot be avoided due to the limitations on data availability it needs to be corrected if possible, in the analysis {@10.2307/20485961, p. 7} this could be achieved through generalization of inverse probability weighting. For this however we would need to categorize developing countries in one group where some have recorded female literacy which then gets multiplied by the number of countries not considered due to missing data. Inverse probability weighting would work as the score for female literacy is not zero for the countries where it is not recorded.  However, a lot of assumptions would need to be made to apply this method. The assumption that the filler-country would be close enough to the unobserved countries is by nature untestable and therefor it is a strong assumption.

@@

$$
MMR=\beta_0 + \beta_1 L+\beta_2 GDP +\beta_3 P+ \beta_4 C
$$
In the formula, P and C respectively account for 3 year periods and the country specific effects.

@@

G-separation calculation has two steps: 
1. Fitting a Q-model, that is, a traditional regression of the outcome on the treatment and  relevant covariates.  In our case, we used the traditional regression fitted in the original paper as Q model.

2. The Q-model is used to predict counterfactual outcomes for all the observations under both treatment settings (L=1 for flexible abortion laws and L=0 if not). More specifically, we plugged both treatment regimes in the the regression fit and obtained two predicted outcomes: $Y_{L=0}=Y_0$ for the unexposed and $Y_{L=1}=Y_1$ for the exposed. Both outcomes are then used to calculate the average $Y_1-Y_0$, which corresponds to the ATE.

@@

However, it is also worth noticing that G-separation also relies on correct model specification, as traditional regression models do, and on positivity and the existence and consistency of counterfactual outcomes (unlike regression). A clear advantage of the ATE estimator obtained by G-separation vis-à-vis regression is that it doesn't estimate the nuissance parameters and focuses only on the (causal) effect of interest.

Furthermore, G-separation allows for the calculation of both ATT and ATU by restricting the ATE estimation to the treated and the untreated respectively[@wang2017g].

@@

*Updated datasets*
Although the datasets used in the original paper were hosted in public websites, the currently available version is not exactly the same. Since both the abortion policies and the maternal mortality sets are now hosted in interactive webpages, it was hard to access the aggregated and historical data from both abortion policies and maternal mortality rates between 1985-2013. Despite these restrictions, we collected data on all the relevant variables from 2000 to 2017. The maternal mortality data was accessed in the World Bank Data repository[@MMR], and the global abortion policies were found in the UN Department for Economic and Social Affairs archive [@WorldAbortionPolicies].

@@

### Regression ATE
In the first part of our exploration, we followed the methodology of the original paper and used fixed effects linear regression models to estimate the relationship between abortion law flexibility and MMR. We controlled for GDP per capita, time in three-year intervals, and country-specific effects through a factor country-name variable. According to our results, flexible abortion laws reduce maternal deaths by 57 per 100.000 live births [ADD TABLE]. This result is close enough to the original result, which found that maternal deaths reduce by 45 per 100,000 live births (95% CI: -64, -26) with flexible abortion laws. 
We also conducted added "female primary education completion" as a control variable, and found very similar result: the effect of abortion law flexibility on maternal mortality appears to be a reduction in the mortality rate of 53 deaths per 100.000 live births.

*Heterogeinity of effects*

@@

*G-separation to find ATE, ATU, ATT*
After following the methodology outlined in @snowden2011implementation, we got an $ATE=-56,9$, which is roughly the same estimation of the effect of policy flexibility on MMR as the one resulting from the regression model. This is not surprising since both models estimated the effect as if it where identical across all strata of covariates because no covariate interactions were specified.

We then calculated the treatment effects on the treated and the untreated and got the following results: $ATT=-56.86$ and $ATU=-2.16e^{-12}$, which means that ATT and the ATE are almost the same and the ATU is almost negligible.

The latter would mean that the potential impact of flexibility on abortion policies on MMR among those countries that have very stringent laws is almost non-existent, contradicting one of the central assumptions of the original paper. However, such a conclusion is not completely justified. On the contrary, this results comes with the need to revise the central assumptions under which the ATT and ATU can be calculated with G-separation. As explained by [@wang2017g], the computation of ATT and ATU also calls for consistency. This is, it requires assuming that the observed outcomes don't depend on the means by which the treatment was assigned. As @cole2009consistency, this assumption is not very hard to fullfil in observational studies (because there is no way to manipulate the treatment). In the specific case of our study, there are aspects of the way the laws are adopted (judicial review, constitutional mandate, etc.) that might affect the outcome.

*Unmeasured confounding*
We conducted a sensitivity analysis to get an idea of the strength of unmeasured counfounders in our model. These were not considered in the traditional regression estimates because to avoid bias, regression estimates require the assumption of no unobserved confounders.

We run the analysis for "female primary completion levels" and gdp  our regression models: 
1. The same we used for the g-separation analysis
2. A different regression model including primary completion levels as confounder.

The robustness value of the sensitivity analysis allows us to quantify the amount of residual variance of both the treatment and the outcome that unobserved confounders must explain in order to remove the observed effect. In the first model, the robustness value was 16%, while for the second one it was 21%. This means that in the second model the unobserved confounders need to be "stronger" to wash away the effect of flexibility laws on MMR.

But how strong must they be to completely remove the effect in both models? The following countour plots show that, while gdp doesn't have a strong effect on MMR, in the second model a counfounder three times as strong as primary_completion would remove the causal effect of interest.

```{r, echo=FALSE,out.width="49%",out.height="49%",fig.show='hold', fig.align='center', fig.cap="Figure X. a) Countour plot for the sensitivity analysis of model 2 using primary completion level as benchmark confounder vs. b) Countour plot for the sensitivity analysis of model 1 using gdp as benchmark confounder"}

knitr::include_graphics(c("Figures/primary.sensitivity.png","Figures/primary.sensitivity.png"))
```

@@

@@

Our analysis can be found in the following github link: [Abortion Laws Repository](https://github.com/aduarte56/AbortionLaws)

@@

* Helena Bakic : Empirical estimation,Identification, Heterogeinity of effects, Analysis
* Anna-Lisa Wirth (213286@mds.hertie-schoo.org) : Background, Unmeasured Confounding, Missing Data, Sensitivity Analysis, Analysis and Conclusion 
* Ángela Duarte Pardo : Data Cleaning, G-separation, Methodology, Modelling and Analysis

@@
